A mixture of hidden Markov models to predict the lymphatic spread in head and neck cancer depending on primary tumor location

### 2024-10-14 _tex/manuscript.tex
Built site for gh-pages

\raggedbottom
\usepackage{multirow}
\usepackage{centernot}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

Head and Neck Squamous Cell Carcinoma (HNSCC) are known to spread
through the lymphatic system often leading to metastases in the lymph
nodes \citep[@shah\_patterns\_1990]{mukherji_cervical_2001}. To minimize
nodal recurrences, lymph node levels (LNL) at risk of harboring occult
metastases are typically irradiated electively. Current guidelines for
different tumor locations are based on the overall prevalence of nodal
disease as reported in literature
\citep[@mukherji\_cervical\_2001, @shah\_patterns\_1990]{biau_selection_2019}.

To personalize the prediction of the risk of occult metastases, given a
patient's individual diagnosis, we previously published a large,
multi-centric dataset where the lymphatic involvement per LNL is
available for each
patient\citep[@ludwig\_multi-centric\_2023]{ludwig_dataset_2022}.Building
on this dataset, we introduced an interpretable hidden Markov model
(HMM), trained to predict the risk for occult nodal disease, given an
individual patient's diagnosis \citep{ludwig_hidden_2021}.

@@

Initially, separate models were trained for distinct tumor locations,
such as the oropharynx and oral cavity. These tumor locations are also
used in guidelines to define the elective target volumes
\citep{biau_selection_2019}. However, this approach did not account for
variations in lymphatic spread between subsites within these tumor
regions. With data for more than 2700 patients available, we can now
further analyze subsite specific spread patterns. Closer analysis showed
that pooling subsites into a single model led to inaccurate predictions,
as it failed to capture distinct lymphatic spread patterns. To resolve
this, we propose using a mixture of HMMs, which allows us to model the
lymphatic spread more accurately for tumors located near anatomical
borders, such as those between the oropharynx and oral cavity (e.g.,
tumors in the palate).

Additionally, we extend the analysis to a broader mixture model that
encompasses tumors of the oral cavity, oropharynx, hypopharynx, and
larynx, offering a more comprehensive understanding of lymphatic spread
across these regions.

@@

The datasets 1-4 are publicly available as CSV tables
(\citet{ludwig_multi-centric_2023} \citet{ludwig_detailed_2022}) and can
be interactively explored on \href{https://lyprox.org}{LyProX}. For each
patient the primary tumor subsite is reportd and each indicidual LNL is
reported as either metastatic or healthy given the available diagnostic
modalities, which include pathology after neck dissection in some
patients. In this work we will stratify the tumor locations into
different ICD codes which are depicted in figure~\ref{fig-subsites}.

@@

The HMM describes each LNL \(v \in 1, 2, ..., V\) by a binary random
variable corresponding to the status of the LNL; healthy (0) or involved
(1). The entire state of a patient with \(V\) LNLs is defined by the
\(v\)-dimensional vector \(\mathbf{X} = [X_1, X_2, ...,X_V]\). In the
HMM, a patient's involvement is modeled over time \(t\). Thus, a
patient's state of lymph node involvement \(\mathbf{X}[t]\) evolves over
discrete time steps \(t\). Let us enumerate all \(2^V\) possible states,
representing all combinations of LNL involvement. In this paper, we
consider ipsilateral LNLs I, II, III, IV and V, which amounts to 32
possible states. The HMM is then specified by a transition matrix
\(\mathbf{A}\):

@@

\begin{equation}\phantomsection\label{eq-likelihood-HMM}{
P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_k^K \left[P \left( \mathbf{X}=\boldsymbol{\xi}_k \mid \boldsymbol{\theta}, T_k \right) \right]
}\end{equation}

In equation~\ref{eq-likelihood-HMM} we compute the likelihood of
observing patient the diagnosis of each patient \(k\), i.e.~t-stage
\(T_k\) and involvement \(\xi_k\).

@@

The mixture model assumes that the data is generated by a set of \(M\)
different lymphatic spread models. Each patient \(k\) is then assumed to
be generated by one specific model from this set of \(M\) models.

To formalize this, each patient \(k\) can be assigned to a specific
model via a binary latent vector \(\boldsymbol{\epsilon}_k\). The vector
\(\boldsymbol{\epsilon}_k\) has length \(M\), with exactly one element
set to 1, indicating which model generated the patient's data, and all
other elements set to 0. Thus, for patient \(k\), the latent variable
\(\boldsymbol{\epsilon}_k\) can be interpreted as a categorical
indicator variable that encodes the assignment to one of the \(M\)
lymphatic spread models.

Next, we introduce subsite-specific mixing parameters
\(\boldsymbol{\pi}^s = \{\pi_1^s, \pi_2^s, \dots, \pi_M^s\}\), where
each \(\pi_m^s\) represents the prior probability that a patient with
tumors in subsite \(s\) is generated by model \(m\). These mixing
parameters must satisfy the condition:

@@

The mixing parameters \(\boldsymbol{\pi}^s\) determine the proportion of
patients with tumors in subsite \(s\) that are expected to be generated
by each model. Specifically, \(\pi_m^s\) is the probability that a
randomly chosen patient with a tumor in subsite \(s\) is generated by
model \(m\).

The latent variables \(\boldsymbol{\epsilon}_{k}\), on the other hand,
are patient-specific and represent the assignment of each patient to a
particular model. The latent variables are central to the model, as they
are hidden (unobserved) and must be inferred.

The joint probability of the observed data \(\mathbf{D}\) (i.e., patient
data) and the latent variables \(\boldsymbol{\epsilon}\) is given by:

\begin{equation}\phantomsection\label{eq-complete-likelihood}{
P(\mathbf{D}, \boldsymbol{\epsilon} | \boldsymbol{\theta}, \boldsymbol{\pi}) = \prod_{k}^{K} \prod_{m}^{M} \left[ \pi_m^{s_k} P(\mathbf{D}_k | \boldsymbol{\theta}_m, \boldsymbol{\pi}) \right]^{\epsilon_{k}^m}
}\end{equation}

@@

\begin{itemize}
\tightlist
\item
  \(\pi_m^{s_k}\) is the mixing parameter for patient \(k\) with a
  primary tumor in subsite \(s_k\) and model \(m\),
\item
  \(P(\mathbf{D}_k | \boldsymbol{\theta}_m)\) is the likelihood of
  patient \(k\)'s data given that it was generated by model \(m\),
\item
  \(\boldsymbol{\theta}_m\) represents the parameters of model \(m\),
\item
  \(\epsilon_{k}^m\) is the latent variable that indicates whether
  patient \(k\) was generated by model \(m\).
\end{itemize}

\textbf{Incomplete data likelihood which can probably be skipped}

Since the latent variables \(\boldsymbol{\epsilon}_k\) are unobserved,
the incomplete complete data likelihood marginalizes over all possible
latent assignments to reflect the uncertainty about which model
generated each patient's data:

\[
P(\mathbf{D} | \boldsymbol{\theta}, \boldsymbol{\pi}) = \sum_{\boldsymbol{\epsilon}} P(\mathbf{D}, \boldsymbol{\epsilon} | \boldsymbol{\theta}, \boldsymbol{\pi})
\]

This allows us to integrate out the latent variables
\(\boldsymbol{\epsilon}\), as their true values are unknown.

\textbf{Continues from here}

The latent variables \(\boldsymbol{\epsilon}_k\) are unknown and must be
inferred from the observed data. The mixture model assumes that each
patient is generated by exactly one of the \(M\) models. To do so a
probabilistic distribution over the latent variables, given the observed
data and model parameters is computed, the posterior distribution:

\[
P(\boldsymbol{\epsilon}_k | \mathbf{D}_k, \boldsymbol{\theta}, \boldsymbol{\pi})
\]

describes the probability that patient \(k\) was generated by each of
the \(M\) models, allowing for soft assignments where each patient is
associated with multiple models with certain probabilities, as opposed
to being strictly assigned to just one. Given the probability
distribution for the latent variables \(\boldsymbol{\epsilon}_k\), we
compute the expectation value
\(\mathbb{E}[\boldsymbol{\epsilon}_k] = \boldsymbol{\gamma}(\boldsymbol{\epsilon}_k)\)
for each patient to then apply the complete data likelihood given in
equation~\ref{eq-complete-likelihood}.

The likelihood of the complete dataset is then expressed as:

\begin{equation}\phantomsection\label{eq-complete-likelihood-explicite}{
P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_k^K \left[ \prod_{m}^M \pi_m^{s_k} P_m \left( \mathbf{X}=\boldsymbol{\xi}_k \mid \boldsymbol{\theta}_m, T_k \right) \right]^{\gamma(\epsilon_k^m)} 
}\end{equation}

In equation~\ref{eq-complete-likelihood} we compute the likelihood of
observing the patient's state \(\boldsymbol{\xi}_k\) with t-stage
\(T_k\) under model \(m\) similarly to equation~\ref{eq-likelihood-HMM}.
However, we weight each model with the mixing parameter \(\pi_m^{s_k}\).
Since the mixing parameters are subsite dependent, we differ between
each subsite.

It is important to mention, that the latent variables
\(\boldsymbol{\epsilon}\) are only used for parameter estimation. The
subsite specific mixture parameters \(\boldsymbol{\pi}^s\) are the only
parameters which are used for later risk predictions. In
figure~\ref{fig-model-simple} the mixture parameters are illustrated.
Subsites with different spread patterns, such as Gum (C03) and Base of
tongue (C01), are expected to get different model assignments.
Nonetheless, the latent variables for two patients with the same
diagnosis, i.e.~same involvement \(\boldsymbol{\xi}\) and and t-stage
\(T\), but different subsites are the same.

@@

The model now contains two sets of parameters: the tumor spread
probabilities \(\mathbf{\theta_m}\) for each model, and the mixing
coefficients \(\pi_m^s\). The posterior distribution over these
parameters \(P(\boldsymbol{\theta},\boldsymbol{\pi}|\mathbf{D})\) can be
estimated assuming a uniform prior on the interval {[}0,1{]} for all
parameters. However,
\(P(\boldsymbol{\theta},\boldsymbol{\pi}|\mathbf{D})\) is a
multimodal-distribution as one can permute the different models and
achieve the same result. To address this problem, we apply an
expectation-maximization (EM) algorithm.

@@

To solve for the model parameters, we apply the EM algorithm, which
iterates between two steps: the \textbf{Expectation (E) step} and
\textbf{the Maximization (M) step}.

\paragraph{Expectation Step}\label{expectation-step}

In the E-step, we compute the expected value of the latent variables
\(\boldsymbol{\epsilon}\), which represent the probabilities of a
patient \(k\) originating from one of the models \(m\). Given the
current estimates of \(\boldsymbol{\theta}\) and \(\boldsymbol{\pi}\),
the expectation value \(\gamma(\epsilon_k^m)\) is calculated as:

\begin{equation}\phantomsection\label{eq-expectation-step}{
\mathbb{E}[\epsilon^m_k] = \gamma(\epsilon^m_{k}) = \frac{\pi_m^{s_k} P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_m}, T_k)}{\sum_{n}^M \pi_n^{s_k} P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_n}, T_k)}
}\end{equation}

\paragraph{Maximization Step}\label{maximization-step}

In the M-step, the model parameters are updated by maximizing the
complete data likelihood in equation
equation~\ref{eq-complete-likelihood-explicite}. For better handling, we
are optimizing the log likelihood:

\begin{equation}\phantomsection\label{eq-maximization-step}{
\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}, \boldsymbol{\pi}) = \sum_k^K \sum_{m}^M \gamma(\epsilon^m_{k}) \left( \ln \pi_m^{s_k} + \ln P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_m}, T_k) \right)
}\end{equation}

In this step, the mixture parameters \(\pi_m^s\) and the model
parameters \(\boldsymbol{\theta}_m\) can be optimized separately. The
mixture parameters \(\pi_m^s\) can be updated analytically:

\begin{equation}\phantomsection\label{eq-mixture-maximization}{
\pi_m^s = \frac{1}{N^s} \sum_k^{N^s} \gamma(\epsilon^m_k)
}\end{equation}

On the other hand, the model parameters \(\boldsymbol{\theta_m}\) are
optimized component-wise using an optimization algorithm, as an analytic
solution is not feasible. The updated log-likelihood for each model is:

\[
\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}_m) = \sum_{k}^K \gamma(\epsilon^m_{k}) \ln P(\mathbf{X} = \boldsymbol{\xi}_i \mid \theta_m, T_k)
\]

By iterating these steps, the EM algorithm converges to a (local)
maximum of the complete data likelihood
equation~\ref{eq-complete-likelihood}.

>>>>

\raggedbottom
\usepackage{multirow}
\usepackage{centernot}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

Head and neck squamous cell carcinomas (HNSCCs) are known to spread
through the lymphatic system often leading to metastases in the lymph
nodes \citep{mukherji_cervical_2001, shah_patterns_1990}. To minimize
nodal recurrences, lymph node levels (LNLs) at risk of harboring occult
metastases are typically irradiated electively. Current guidelines for
different tumor locations are based on the overall prevalence of nodal
disease as reported in literature
\citep{biau_selection_2019, mukherji_cervical_2001, shah_patterns_1990}.

To personalize the prediction of the risk of occult metastases, given a
patient's individual diagnosis, we previously published a large,
multi-centric dataset where the lymphatic involvement per LNL is
available for each
patient\citep{ludwig_dataset_2022, ludwig_multi-centric_2023}. Building
on this dataset, we introduced an interpretable hidden Markov model
(HMM), trained to predict the risk for occult nodal disease, given an
individual patient's diagnosis \citep{ludwig_hidden_2021}.

@@

Initially, separate models were trained for distinct tumor locations,
such as the oropharynx and oral cavity. These tumor locations are also
used in guidelines to define the elective target volumes
\citep{biau_selection_2019}. However, this approach did not account for
variations in lymphatic spread between subsites within these tumor
regions. With data from more than 2700 patients available, we can now
further analyze subsite specific spread patterns. Closer analysis showed
that pooling subsites into a single model led to inaccurate predictions,
as it failed to capture distinct lymphatic spread patterns. To resolve
this, we propose using a mixture of HMMs, which allows us to model the
lymphatic spread more accurately for tumors located near anatomical
borders, such as those between the oropharynx and oral cavity (e.g.,
tumors in the palate).

Additionally, we extend the analysis to a broader mixture model that
encompasses tumors of the oral cavity, oropharynx, hypopharynx, and
larynx, resulting in further personalized predictions of lymphatic
spread across these regions.

@@

The datasets 1-4 are publicly available as CSV tables
\citep{ludwig_multi-centric_2023, ludwig_detailed_2022} and can be
interactively explored on \href{https://lyprox.org}{LyProX}. For each
patient the primary tumor subsite is reportd and each indicidual LNL is
reported as either metastatic or healthy given the available diagnostic
modalities, which include pathology after neck dissection in some
patients. In this work we will stratify the tumor locations into
different ICD codes which are depicted in figure~\ref{fig-subsites}.

@@

The HMM describes each LNL \(v \in 1, 2, ..., V\) by a binary random
variable corresponding to the status of the LNL; healthy (0) or involved
(1). The entire state of a patient with \(V\) LNLs is defined by the
\(V\)-dimensional vector \(\mathbf{X} = [X_1, X_2, ...,X_V]\). In the
HMM, a patient's involvement is modeled over time \(t\). Thus, a
patient's state of lymph node involvement \(\mathbf{X}[t]\) evolves over
discrete time steps \(t\). Let us enumerate all \(2^V\) possible states,
representing all combinations of LNL involvement. In this paper, we
consider ipsilateral LNLs I, II, III, IV and V, which amounts to 32
possible states. The HMM is then specified by a transition matrix
\(\mathbf{A}\):

@@

\begin{equation}\phantomsection\label{eq-likelihood-HMM}{
P \left( \mathbf{D} \mid \boldsymbol{\theta} \right) = \prod_k^K P \left( \mathbf{X}_k=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}, T_k \right)
}\end{equation}

In equation~\ref{eq-likelihood-HMM} we compute the likelihood of
observing patient the diagnosis of each patient \(k\), i.e.~t-stage
\(T_k\) and involvement \(\mathbf{X}_k=\boldsymbol{\xi}_i\).

@@

The mixture model assumes that the data is generated by a set of \(M\)
different lymphatic spread models. Each patient \(k\), with their
primary tumor in subsite \(s \in (1, 2, \ldots, S)\), is assumed to be
generated by one specific model \(m \in (1, 2, \ldots, M)\) from this
set of \(M\) models with probability \(\pi_m^s\). These so-called
\emph{mixing parameters}
\(\boldsymbol{\pi}^s = \{\pi_1^s, \pi_2^s, \dots, \pi_M^s\}\) must
satisfy the condition

@@

If we could record the componet \(m\) from which a patient \(k\) was
drawn from, we could store this information in a binary latent vector
\(\boldsymbol{\epsilon}_k\). The vector \(\boldsymbol{\epsilon}_k\) has
length \(M\), with exactly the \(m\)-th element set to 1, indicating
which model generated the patient's data, and all other elements set to
0. Thus, for patient \(k\), the latent variable
\(\boldsymbol{\epsilon}_k\) can be interpreted as a categorical
indicator variable that encodes the assignment to one of the \(M\)
lymphatic spread models. Typically in mixture models, this so-called
\emph{latent variable} is unknown. However, it can be inferred and is
useful for inferring the models' parameters.

The joint probability of the observed data \(\mathbf{D}\) (i.e., patient
data) and the latent variables \(\boldsymbol{\epsilon}\) - sometimes
called the \emph{complete data likelihood} - is given by:

\begin{equation}\phantomsection\label{eq-complete-likelihood}{
P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}, \boldsymbol{\pi}) = \prod_{k}^{K} \prod_{m}^{M} \left[ \pi_m^{s_k} P(\mathbf{D}_k \mid \boldsymbol{\theta}_m) \right]^{\epsilon_{k}^m}
}\end{equation}

@@

\begin{itemize}
\tightlist
\item
  \(\pi_m^{s_k}\) is the mixing coefficient for subsite \(s_k\) (where
  patient \(k\) has their tumor) and model \(m\),
\item
  \(P(\mathbf{D}_k \mid \boldsymbol{\theta}_m)\) is the likelihood of
  patient \(k\)'s diagnosis, given that it was generated by model \(m\),
\item
  \(\boldsymbol{\theta}_m\) represents the parameters of model \(m\),
\item
  \(\epsilon_{k}^m\) is the latent variable that indicates patient \(k\)
  was generated by model \(m\).
\end{itemize}

\begin{tcolorbox}[enhanced jigsaw, breakable, colbacktitle=quarto-callout-note-color!10!white, left=2mm, colback=white, coltitle=black, arc=.35mm, bottomrule=.15mm, rightrule=.15mm, bottomtitle=1mm, opacitybacktitle=0.6, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{TODO: Discuss if we need this}, titlerule=0mm, toptitle=1mm, toprule=.15mm, colframe=quarto-callout-note-color-frame, leftrule=.75mm, opacityback=0]

In contrast, the \emph{incomplete data likelihood} marginalizes over all
possible latent assignments to reflect the uncertainty about which model
generated each patient's data:

\begin{equation}\phantomsection\label{eq-incomplete-likelihood}{
P(\mathbf{D}, \mid \boldsymbol{\theta}, \boldsymbol{\pi}) = \prod_{k}^{K} \sum_{m}^{M} \pi_m^{s_k} P(\mathbf{D}_k \mid \boldsymbol{\theta}_m)
}\end{equation}

Ultimately, we want to find the parameters which maximize this
likelihood function for the given data.

Note the summation inside the product. This structure of mixture models
makes naive inference difficult, because the logarithm of this quantity
is expensive to compute and not easy to differentiate. Thus, inferring
the latent assignment is helpful, because the complete data
(log-)likelihood (equation~\ref{eq-complete-likelihood}) does not suffer
from this shortcoming.

\end{tcolorbox}

To infer the latent assignment \(\boldsymbol{\epsilon}_k\), we start
with its distribution, given the observed data and model parameters:

\begin{equation}\phantomsection\label{eq-responsibilities}{
\gamma(\epsilon_k^m) := \mathbb{E}[\epsilon_k^m] = \frac{\pi_m^{s_k} P \left( \mathbf{D}_k \mid \boldsymbol{\theta}_m \right)}{\sum_{j \leq M} \pi_j^{s_k} P \left( \mathbf{D}_k \mid \boldsymbol{\theta}_j \right)}
}\end{equation}

This expectation value - often called the \emph{responsibility} -
describes the probability that patient \(k\) was generated by model
\(m\). It can be used to compute the expected complete data likelihood:

\begin{equation}\phantomsection\label{eq-expected-complete-likelihood}{
\mathbb{E}_{\boldsymbol{\epsilon}} \left[ P \left( \mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}, \boldsymbol{\pi} \right) \right] = \prod_{k=1}^K \prod_{m=1}^M \left[ \pi_m^{s_k} P \left( \mathbf{D}_k \mid \boldsymbol{\theta}_m \right) \right]^{\gamma(\epsilon_k^m)}
}\end{equation}

This expected complete data likelihood has the same tractable form as
equation~\ref{eq-complete-likelihood} and thus allows us to infer both
the mixing coefficients \(\boldsymbol{\pi}\) as well as each component
model's parameters \(\boldsymbol{\theta}_m\) using straightformward
inference. Also, these two sets of parameters are the only ones used for
the later risk prediction. The responsiblities \(\gamma(\epsilon_k^m)\)
are only used during inference.

In figure~\ref{fig-model-simple} the mixture coefficients are
illustrated. Subsites with different spread patterns, such as Gum (C03)
and Base of tongue (C01), are expected to get different model
assignments. Nonetheless, the latent variables for two patients with the
same diagnosis, i.e.~same involvement \(\boldsymbol{\xi}\) and and
t-stage \(T\), but different subsites are the same.

@@

@@

To actually find the mixing coefficients \(\boldsymbol{\pi}\) and all
models' parameters \(\boldsymbol{\theta}_m\) that maximize
equation~\ref{eq-incomplete-likelihood}, we follow an iterative approach
called \emph{expectation-maximization} or \emph{EM-algorithm}. With
arbitrarily initialized starting parameters, we alternate between the
following two steps:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  In the \textbf{E}xpectation step, we compute the responsibilities
  \(\gamma(\epsilon_k^m)\), which represent the probabilities of a
  patient \(k\) originating from one of the models \(m\), given the
  current estimates of \(\boldsymbol{\theta}\) and \(\boldsymbol{\pi}\),
  as given in equation~\ref{eq-responsibilities}.
\item
  During the \textbf{M}maximization step, we find a new set of
  parameters that maximize the new expected complete data likelihood
  (equation~\ref{eq-expected-complete-likelihood}). For the mixture
  coefficients, we can even find an analytic solution to the new
  maximum: \[
  \pi_m^s = \frac{1}{|K_s|} \sum_{k \in K_s} \gamma(\epsilon_k^m)
  \] where we sum over the set of all patients \(K_s\) with their tumor
  in subsite \(s\).\\
  The models' new parameters are found by numerically maximizing the
  respective likelihood, weighted by the responsibilities:
  \begin{equation}\phantomsection\label{eq-maximization-step}{
  \ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}_m) = \sum_k^K \gamma(\epsilon^m_{k}) \big[ \ln \pi_m^{s_k} + \ln P \left( \mathbf{D}_k \mid \boldsymbol{\theta}_m \right) \big]
  }\end{equation}
\end{enumerate}

By iterating these steps, the EM algorithm is guaranteed to converge to
a (local) maximum of the complete data likelihood
(equation~\ref{eq-complete-likelihood}).
